Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-13 manuscript.tex
Removed comments from the paper

\todo[inline]{This is an example of how to leave a comment in a Latex
document.}

@@

Previous work by \citet{Ku1977} investigated the use of the GLQ in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criteria that the distance
between adjacent point masses should be less than the distance to the computation point.
\citet{Asgharzadeh2007} used  this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, an analysis of how well this ad hoc criteria of
\citet{Ku1977} works for gravity gradient components or for tesseroids
has never been done before.
There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.
\todo[inline]{LEO: Valeria, I've made some modifications to your correction.}

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.
Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.
This division effectively increases the total number of point masses
because the number of points masses per tesseroid is fixed.
A tesseroid is divided if the minimum distance to the computation point
is smaller than the largest dimension of the tesseroid.
\todo[inline]{VAL:
Leo, I have modified the above sentence because I have not understood the original meaning.
}
\todo[inline]{LEO: There was a mistake in that sentence.
I've rewritten it. Is it better now?}
This division is repeated recursively until all tesseroids obey the criterion.
Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.
The advantage of this adaptive discretization over
increasing the number of points masses is that the
total distribution of point masses will be greater
only close to the computation point.
This makes the adaptive discretization more computationally efficient.

@@

\todo[inline]{Valeria: Leo,  what does the above sentence mean? Does it mean that these mentioned authors have already used a former version of this software?}
\todo[inline]{Leo: exactly. These are the papers I found that mentioned using Tesseroids.}

@@

The advantage of this adaptive discretization is
that the number of point masses is only increased
in parts of the tesseroid that are
closer to the computation point.
Notice that the alternative approach of
simply increasing the order of the GLQ
would increase the number of point masses
evenly throughout the whole tesseroid.
\todo[inline]{LEO: Valeria, this correction you made is wrong.
It is not ``in regions where the tesseroids are closer''.
It is actually ``in parts of the tesseroid that are closer''.
This is all happening inside a single tesseroid.
I have corrected the text above.}

@@

\emph{Tesseroids} consists of command-line programs
written in the C programming language.
The package includes programs to calculate
the gravitational fields of tesseroids and
rectangular prisms (in both Cartesian and spherical coordinates).
\todo[inline]{Valeria: Leo, I did not understand "as well as model and computation grid generation"}
\todo[inline]{Leo: I mean that there are programs to generate the model files and the grid files. This can be removed without any problem.}
All programs receive input through
command-line arguments and the standard input channel (``STDIN'')
and output the results through the standard output channel (``STDOUT'').
For example,
the command to generate a regular grid with $NLON \times NLAT$ points,
calculate on it $g_z$ and $g_{zz}$ caused by
the tesseroids in a file ``MODELFILE'',
and save the results to a file called ``OUTPUT''
is:

@@

\begin{equation}
    \frac{d}{L_i} > D,
    \label{eq:condition}
\end{equation}
\todo[inline]{Leo: the condition was wrong. It had a $<$ when it should be a $>$. I fixed it.}

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last tesseroid from it).
        This will cause \emph{stktop} to be reduced by one.
        This tesseroid is the one that will be evaluated in the following
        steps.
    \item Compute the distance $d$ (equation~\ref{eq:distance}) between
        the geometric center of the tesseroid and
        the computation point.
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ using equations~\ref{eq:sizelon}-\ref{eq:sizer}.
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.

\todo[inline]{  Regarding the Item 5  - I think that it is confusing.

In the item 5 is written " If none of the dimensions fail the above condition, ...."

This  means that all the dimensions hold the inequality (21). Ok ?

So, Item 5 could be written directly,

5. If all the dimensions hold the inequality (21), hence the tesseroid is not divided and the gravitational effect of the tesseroid is computed by using the
        Gauss-Legendre Quadrature.....}
\todo[inline]{Leo: I agree completely. I have made the modifications below.}

    \item If all dimensions hold the inequality \ref{eq:condition},
        the tesseroid is not divided and its
        gravitational effect is computed using the
        Gauss-Legendre Quadrature
        (equations~\ref{eq:tesspot}-\ref{eq:tesstensor} and~\ref{eq:glq3d}).
        We use a GLQ order of two for all three dimensions
        ($2 \times 2 \times 2 = 8$ point masses)
        by default.
        This value can be changed using
        a command-line argument of the modeling programs.
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.

\todo[inline]{   Regarding the Item 6b  - I am not sure, but I think that it might be wrong or written in an awkward style

Item 6b is written - "Check if the number of new elements plus \emph{stktop} is smaller than the maximum stack size."
Based on item 6c, I think that in 6b you are check if the stack is full, ok ?

My suggestion is to merge items 6b and 6c, for example:

If the stack is full, place the smaller tesseroids into the stack; otherwise warn the user of "stak overflow" and compute the effect of the larger tesseroid (step 5)

}
\todo[inline]{Leo: Yes, that sentence was a bit confusing.
The idea is checking if the newly generated tesseroids
fit in the stack.
I just wanted to explain how to check for that
(the number of new elements ...).
I'll rephrase it.}
\todo[inline]{Leo: By the way, the proper way of writing " in LaTeX is ``like this''.}

        \item Check if there is room in the stack
            for the new tesseroids
            (i.e.,the number of new elements plus \emph{stktop}
             is smaller than the maximum stack size).
             If there isn't,
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
             If there is room in the stack,
             place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\emph{stktop} is equal to -1).
\end{enumerate}

@@

\todo[inline]{ VAL: The code was written in C., was not it? I think that there is a  version in Python. However, the readers do not known. }

\todo[inline]{Leo: The code that we are discussing in this paper was written in C.
The figures were generated in IPython notebooks in Python.
The error analysis was done in IPython notebooks as well.
However, it does NOT use the Python version of the tesseroid code.
It runs the C programs and uses Python to calculate the differences
The modeling that is written in Python is not included or discussed here.
Readers do not need to know that it exists.}

\todo[inline]{VAL: The above sentence is confusing:
1) "These files combine the analysis and figure.... ",
What are the files? Probably, they are the IPython notebooks. Ok?
Why are you saying "the analysis"
In the last paragraph, you said "running the analysis".
What is the analysis? Is the test to evaluate the accuracy?
I think that you want to say
"The Ipython nootebook files combine source codes in python, code executions, text, equations, dnd the figures generated by the code into a single document"
You need keeping in mind that you know about Ipython nootebook. However, the readers cannot know.}

\todo[inline]{LEO: Yes, that part is a bit confusing now that I read it again.
By analysis, I mean the error analysis.
That is done in \href{http://nbviewer.ipython.org/github/pinga-lab/paper-tesseroids/blob/master/notebooks/tesseroid_vs_spherical_shell.ipynb}{this IPython notebook}.
This file will be  available  with the article.
There is a link to this online version in \url{https://github.com/pinga-lab/paper-tesseroids}
which is mentioned in the text.
There are two other notebooks that make the figures for the methods and the final results.
There are links to those two in the above link  as well.
I'll rewrite the paragraph to make it more clear.}

@@

\todo[inline]{ VAL: The above paragraph is great !}

@@

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points at different geographic locations
(see Table~\ref{tbl:experiment}).
\todo[inline]{ VAL: A grid of 10 X  10  point covers a are of ??? X ???? km}
\todo[inline]{LEO: The grid location varied in each experiment of the numerical analysis. Table 1 shows the locations of the grids.
This is discussed below.
I'll add this information to the text.}
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid.
However, we will consider only the largest error in our analysis.

\todo[inline]{ VAL: I rewrite the above sentence by:
"We compute the differences between the effects of the shell and the tesseroid model on the grid; however, we consider  the largest error only.  }
\todo[inline]{LEO: Rewritten. Is it better now?}
We placed the grid on top of a particular tesseroid
to increase the chances of capturing the true largest integration error.
\todo[inline]{ VAL:I rewrite the above sentence as:
"We decided for the nonconservative procedure of placing the grid on top of a particular tesseroid to increase the chances
of capturing the true largest integration error"}
\todo[inline]{LEO: Changed. I don't think this is really
``nonconservative'' so I left it out.}

We calculate the errors for values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.
Furthermore, we repeated the error analysis in four different numerical experiments,
each with computation grids at different locations and
different tesseroid model sizes.
Table~\ref{tbl:experiment} describes the different numerical
experiments and the corresponding parameters of the computation grid and
tesseroid model.

 \todo[inline]{ Val: Regarding the visual aspect of Figure 4. I think that  the dashed lines are yielding a visual pollution. However, they are helping us to identify the values in the axes. Regarding the visual aspect of Figure 4.
A possibility would be keep the dashed lines, but with  other color (e.g., light gray  dashed lines) }

@@

\todo[inline]{ VAL: I think that we discussed that similar results that are achieved in gzz (Figure 5) also are I achieved in gxx, gyy.... I think that this information could be written.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/distance-size-curves}
    %\missingfigure{}
    \caption{
        The maximum difference between the computed tesseroid and shell effects
        as a function of the distance-size ratio $D$
        for (a) the gravitational potential, (b) $g_z$, and (c) $g_{zz}$.
        The difference is given as a percentage of the shell effect.
        Curves correspond to the different tesseroid models and computation
        grids shown in Table~\ref{tbl:experiment}.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
    }
    \label{fig:dist-size-curves}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/gzz-with-height}
    %\missingfigure{}
    \caption{
        Difference between the computed $g_{zz}$ for the spherical shell and
        the tesseroid model at different heights. Curves show the maximum
        difference as a percentage of the shell value.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
    }
    \label{fig:gzz-with-height}
\end{figure}

@@

\todo[inline]{ I am here. I did not read the captions of the Figures 4 and 5 and the caption of the Table 1}

>>>>

@@

Previous work by \citet{Ku1977} investigated the use of the GLQ
in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criteria that the distance
between adjacent point masses should be less than
the distance to the computation point.
\citet{Asgharzadeh2007} used  this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, an analysis of how well this ad hoc criteria of
\citet{Ku1977} works for gravity gradient components or for tesseroids
has never been done before.
There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.
Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.
This division effectively increases the total number of point masses
because the number of points masses per tesseroid is fixed.
A tesseroid is divided if the minimum distance to the computation point
is smaller than the largest dimension of the tesseroid.
This division is repeated recursively until all tesseroids obey the criterion.
Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.
The advantage of this adaptive discretization over
increasing the number of points masses is that the
total distribution of point masses will be greater
only close to the computation point.
This makes the adaptive discretization more computationally efficient.

@@

@@

The advantage of this adaptive discretization is
that the number of point masses is only increased
in parts of the tesseroid that are
closer to the computation point.
Notice that the alternative approach of
simply increasing the order of the GLQ
would increase the number of point masses
evenly throughout the whole tesseroid.

@@

\emph{Tesseroids} consists of command-line programs
written in the C programming language.
The package includes programs to calculate
the gravitational fields of tesseroids and
rectangular prisms (in both Cartesian and spherical coordinates).
All programs receive input through
command-line arguments and the standard input channel (``STDIN'')
and output the results through the standard output channel (``STDOUT'').
For example,
the command to generate a regular grid with $NLON \times NLAT$ points,
calculate on it $g_z$ and $g_{zz}$ caused by
the tesseroids in a file ``MODELFILE'',
and save the results to a file called ``OUTPUT''
is:

@@

\begin{equation}
    \frac{d}{L_i} > D,
    \label{eq:condition}
\end{equation}

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last tesseroid from it).
        This will cause \emph{stktop} to be reduced by one.
        This tesseroid is the one that will be evaluated in the following
        steps.
    \item Compute the distance $d$ (equation~\ref{eq:distance}) between
        the geometric center of the tesseroid and
        the computation point.
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ using equations~\ref{eq:sizelon}-\ref{eq:sizer}.
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If all dimensions hold the inequality \ref{eq:condition},
        the tesseroid is not divided and its
        gravitational effect is computed using the
        Gauss-Legendre Quadrature
        (equations~\ref{eq:tesspot}-\ref{eq:tesstensor} and~\ref{eq:glq3d}).
        We use a GLQ order of two for all three dimensions
        ($2 \times 2 \times 2 = 8$ point masses)
        by default.
        This value can be changed using
        a command-line argument of the modeling programs.
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if there is room in the stack
            for the new tesseroids
            (i.e.,the number of new elements plus \emph{stktop}
             is smaller than the maximum stack size).
             If there isn't,
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
             If there is room in the stack,
             place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\emph{stktop} is equal to -1).
\end{enumerate}

@@

@@

@@

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points at different geographic locations
(see Table~\ref{tbl:experiment}).
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid.
However, we will consider only the largest error in our analysis.

We placed the grid on top of a particular tesseroid
to increase the chances of capturing the true largest integration error.
We calculate the errors for values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.
Furthermore, we repeated the error analysis in four different numerical experiments,
each with computation grids at different locations and
different tesseroid model sizes.
Table~\ref{tbl:experiment} describes the different numerical
experiments and the corresponding parameters of the computation grid and
tesseroid model.

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/distance-size-curves}
    \caption{
        The maximum difference between the computed tesseroid and shell effects
        as a function of the distance-size ratio $D$
        for (a) the gravitational potential, (b) $g_z$, and (c) $g_{zz}$.
        The difference is given as a percentage of the shell effect.
        Curves correspond to the different tesseroid models and computation
        grids shown in Table~\ref{tbl:experiment}.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
    }
    \label{fig:dist-size-curves}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/gzz-with-height}
    \caption{
        Difference between the computed $g_{zz}$ for the spherical shell and
        the tesseroid model at different heights. Curves show the maximum
        difference as a percentage of the shell value.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
    }
    \label{fig:gzz-with-height}
\end{figure}

@@
